---
title: "Explore Swimmers"
subtitle: "Interactive exploration of Olympic swimming medalists (1900–2016)"
format:
  html:
    toc: true
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(plotly)

# Load data and filter for swimming medalists with known dimensions
swimmers <- read_csv("data/olympics.csv.gz", show_col_types = FALSE) |>
  filter(
    sport == "Swimming",
    !is.na(height),
    !is.na(weight),
    !is.na(medal)
  ) |>
  mutate(
    stroke = case_when(
      str_detect(event, "Freestyle") ~ "Freestyle",
      str_detect(event, "Breaststroke") ~ "Breaststroke",
      str_detect(event, "Backstroke") ~ "Backstroke",
      str_detect(event, "Butterfly") ~ "Butterfly",
      str_detect(event, "Medley") ~ "Medley",
      TRUE ~ "Other"
    ),
    medal = factor(medal, levels = c("Gold", "Silver", "Bronze"))
  )
```

## How to Explore

Use the interactive plot below to examine **2,487 Olympic swimming medalists**:

- **Hover** over any point to inspect the athlete's name, Olympic Games, event, medal, height, and weight.
- **Click or double-click** stroke categories in the legend on the right to focus on a single discipline.
- **Click and drag** to zoom into any cluster of interest, and double-click the plot area to reset the view.

```{r}
#| label: interactive-plot
#| echo: false
#| warning: false
#| message: false

# Create ggplot with formatted tooltip text
p <- swimmers |>
  ggplot(aes(
    x = height,
    y = weight,
    color = stroke,
    text = paste0(
      "<b>", name, "</b> (", team, ")<br>",
      "Games: ", year, " ", city, "<br>",
      "Event: ", event, "<br>",
      "Medal: ", medal, "<br>",
      "Height: ", height, " cm | Weight: ", weight, " kg"
    )
  )) +
  geom_point(alpha = 0.65, size = 1.8) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Height and Weight of Olympic Swimming Medalists",
    x = "Height (cm)",
    y = "Weight (kg)",
    color = "Discipline"
  )

# Convert to interactive plotly widget
ggplotly(p, tooltip = "text")
```

## Observations from the Data

- **Sprinters vs. Distance:** Freestyle swimmers span the widest range, with sprint medalists clustering in higher height and weight ranges.
- **Stroke Differences:** Breaststroke and Butterfly athletes exhibit dense clusters around 180–190 cm for men and 170–180 cm for women, reflecting the power demands of short-axis strokes.
