---
title: "Explore Disciplines & Risk Profiles"
subtitle: "Interactive exploration of Olympic athlete physiological profiles across eight major sports"
format:
  html:
    toc: true
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(plotly)

# Sports under study
study_sports <- c(
  "Basketball",
  "Volleyball",
  "Swimming",
  "Rowing",
  "Athletics",
  "Cycling",
  "Gymnastics",
  "Weightlifting"
)

# Load data for Olympic medalists from modern era (1980–2016)
df <- read_csv("data/olympics.csv.gz", show_col_types = FALSE) |>
  filter(
    sport %in% study_sports,
    !is.na(height),
    !is.na(weight),
    !is.na(medal),
    year >= 1980
  ) |>
  mutate(
    bmi = round(weight / (height / 100)^2, 1),
    medal = factor(medal, levels = c("Gold", "Silver", "Bronze"))
  )
```

## How to Use This Explorer

This interactive visualization displays **`r format(nrow(df), big.mark = ",")` Olympic medalists** from 1980 to 2016 across eight key sporting disciplines:

- **Hover** over any point to inspect the athlete's name, country, event, Olympic Games, medal, height, weight, and BMI.
- **Double-click** any sport in the legend on the right to isolate that specific discipline.
- **Single-click** legend entries to toggle individual sports on and off for direct side-by-side comparison (for example, comparing *Cycling* directly against *Weightlifting*).
- **Click and drag** to zoom into any cluster, and double-click the chart background to reset the view.

```{r}
#| label: interactive-scatter
#| echo: false
#| warning: false
#| message: false

p <- df |>
  ggplot(aes(
    x = height,
    y = weight,
    color = sport,
    text = paste0(
      "<b>", name, "</b> (", team, ")<br>",
      "Sport: ", sport, "<br>",
      "Event: ", event, "<br>",
      "Medal: ", medal, " (", year, " ", city, ")<br>",
      "Height: ", height, " cm | Weight: ", weight, " kg<br>",
      "BMI: ", bmi
    )
  )) +
  geom_point(alpha = 0.65, size = 1.8) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Physiological Profiles of Olympic Medalists (1980–2016)",
    x = "Height (cm)",
    y = "Weight (kg)",
    color = "Discipline"
  )

ggplotly(p, tooltip = "text")
```

## Sport Profiles & Associated Testing Risks

The table below summarizes median physical metrics across our sample and highlights the primary pharmacological risk categories identified in anti-doping literature:

| Discipline | Typical Stature | Mechanical Demand | Primary Doping Risk Profile |
|:---|:---:|:---|:---|
| **Weightlifting** | Compact / Heavy | Maximum force generation | Anabolic androgenic steroids (AAS), growth hormone |
| **Gymnastics** | Short / Lean | High strength-to-weight ratio | Diuretics (weight management), stimulants |
| **Cycling** | Moderate / Lean | Aerobic efficiency, power-to-weight | Blood doping, recombinant EPO, blood transfusions |
| **Rowing** | Tall / Heavy | Lever power and aerobic capacity | Anabolic agents, EPO, endurance enhancers |
| **Athletics** | Variable | Sprint power vs. distance endurance | AAS (sprints/throws) vs. EPO/blood manipulation (distance) |
| **Swimming** | Tall / Long levers | Upper-body propulsion, low drag | Anabolic agents, peptide hormones, recovery enhancers |
| **Basketball** | Very Tall | Height leverage, vertical jump | Stimulants, cannabinoids, recovery peptide hormones |
| **Volleyball** | Tall / Lean | Explosive vertical power, lateral quickness | Stimulants, anabolic recovery agents |
